Linear instability underlying synchronous waving of marine grass in flow

### 2014-08-29 Paper/grass.tex
Introduced \epsilon = (\delta/H)^3
Need to systematically replace $\delta/H$ with \epsilon^{1/3} everywhere.

We use a mean field model for the coupling between the flow and the canopy. Drag force exerted by the grass is approximated via a 
continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$ in the fluid momentum balance\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06} as
\begin{equation}
\rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}
\label{navier-stokes}
\end{equation}
%\begin{equation}
% \mathbf{f}=-N\mathbf{f_{d}}
%\end{equation}
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area, $\rho$ the fluid density, $\mathbf{u}$ the velocity, 
$P$ the pressure, $\mu$ the turbulent viscosity and $\mathbf{g}$ the acceleration due to gravity. The Reynolds number of the flow on the scale of a grass blade is expected to 
be O($10^2$-$10^3$), and hence the drag force is dominated by form drag, and is modeled 
as $\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\hat{n}+C_{T}\rho u_{T}^{2}d\hat{t}$ where $d$ is average width of grass blade
$C_{N}$ and $C_{T}$ are normal and tangential drag coefficients respectively which are set to zero outside the grass; $\bu_{T}$, $\bu_{N}$ are velocity vector along and
normal to grass while $\hat{t},\hat{n}$ being unit vector along and normal to grass. We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. In the field, both $C_N \& N_g$ 
vary with distance from the bottom but we do not expect these variation to be central to the mechanism and therefore take $C_N$ \& $N_g$ to be constants. Based on previous experiments\cite{Vivoni98}, we take $C_N \approx 1$.

For consistency of analysis, we calculate the steady flow as a solution of \eqref{navier-stokes}, and examine its linear stability. The steady uniform flow $U(y)$ resulting from solution of \eqref{navier-stokes} is expected to be unstable, leading to coherent eddies underlying monami. Steady state flow driven by constant pressure gradient satisfies
$-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< H$. 
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=H$) of the fluid layer, as can be seen from the comparison 
between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{fig:basicflow}. Flow consist of a region of approximately constant velocity
\small$U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$\normalsize arising from balance of drag force with pressure gradient with-in grass and a parabolic velocity 
profile in unvegetated region due to balance of viscous force and pressure gradient. Continuity of flow speed and shear stress at canopy top 
give rise to a boundary layer in the flow profile near the tip of the grass. The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this boundary layer thickness is obtained by balance of viscous force with drag, and matching the shear with that in unvegetated region. This leads to a shear layer 
thickness $\delta \sim  H\left[({\rho U_0 H}/\mu) (C_N d N_g H)\right]^{-1/3}$ where $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity and
$H$ is half channel width. Its dependence on the grass density $N_g$ gives us a direct control of its thickness. 
\begin{figure}[htb]
  \includegraphics[scale=1]{Grass_Base_Vivoni_shear}
\caption{Base flow profile (left )from the experiment of Vivoni\cite{Vivoni98} (Case-3A (Figure-3.6A) with 330 plants/$m^2$, 6 blades per plant, 
plant height = 16.0$\pm0.2$ and blade width of 0.28cm)
 and it's approximation with $U_0=31cm/s$ and $\delta = 0.2H$ with our model. The model of canopy is shown in the bottom and the variation of shear layer in the inset }
\label{fig:basicflow}
\end{figure}

Knowing the scale of shear layer we can systematically investigate its effect on flow instability by varying grass number density $N_g$ and pressure gradient ${dP}/{dx}$. Considering small perturbations $u, v, p$ about the base profile $U$ and $P$ respectively, momentum and mass balance equation expanded in first order of perturbed variables yields
\begin{equation}
\begin{split}
\frac{\del u}{\del t}+U\frac{\del u}{\del x}+v\frac{\del U}{\del y} &= -\frac{1}{\rho}\frac{\del p}{\del x}+\frac{\mu}{\rho}\nabla^2u-2S(y) C_{N}dN_{g}Uu, \\
\frac{\del v}{\del  t}+ U\frac{\del v}{\del x} &= -\frac{1}{\rho}\frac{\del p}{\del y}+\frac{\mu}{\rho}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0.
 %\nabla\cdot \bu &= 0
\end{split}
\end{equation}
Momentum and mass balance equation can be non-dimensionalized using half channel height $H$, associated advection time H/$U_0$ and velocity $U_0$. 
We are left with three non-dimensional parameters in the model, viz. the Reynolds number $\Rey= {\rho U_0 H}/{\mu}$, the vegetation drag parametrized as $\delta/H$, 
and the submergence ratio of the grass $h_g/H$. With these scaling, and using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a wave solution of the form $\left(u,v,\psi \right)= \left(\hat u, \hat v, \hat\phi \right)e^{ikx+\sigma t}$ to  obtain a modified Orr-Sommerfield equation. The eigenvalue $\sigma$ for a given $k$ provides growth rate and frequency associated with a perturbation of wavelength $2\pi/k$ through the solution of
% \small
\begin{equation}
\begin{split}
\left(\frac{\sigma}{ik}+U\right) \left(D^2-k^2\right)\phi &= \frac{1}{ik\Rey}\left[D^2 -k^{2} \right]^2\phi +U_{yy}\phi \\
&-\frac{\del}{\del y}\left(\frac{2}{ik\Rey}\frac{H^3}{\delta^3}S(y) U\phi_y\right)
\label{Orr-somerfield}
\end{split}
\end{equation}
\shreyas{Boundary conditions?}
% \normalsize

>>>>

We use a mean field model for the coupling between the flow and the canopy. Drag force exerted by the grass is approximated via a 
continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$ in the fluid momentum balance\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06} as
\begin{equation}
\rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}, \quad \grad\cdot\bu = 0,
\label{navier-stokes}
\end{equation}
%\begin{equation}
% \mathbf{f}=-N\mathbf{f_{d}}
%\end{equation}
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area, $\rho$ the fluid density, $\mathbf{u}$ the velocity, 
$P$ the pressure, $\mu$ the turbulent viscosity and $\mathbf{g}$ the acceleration due to gravity. The Reynolds number of the flow on the scale of a grass blade is expected to 
be O($10^2$-$10^3$), and hence the drag force is dominated by form drag, and is modeled 
as $\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\hat{n}+C_{T}\rho u_{T}^{2}d\hat{t}$ where $d$ is average width of grass blade
$C_{N}$ and $C_{T}$ are normal and tangential drag coefficients respectively which are set to zero outside the grass; $\bu_{T}$, $\bu_{N}$ are velocity vector along and
normal to grass while $\hat{t},\hat{n}$ being unit vector along and normal to grass. We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. In the field, both $C_N \& N_g$ 
vary with distance from the bottom but we do not expect these variation to be central to the mechanism and therefore take $C_N$ \& $N_g$ to be constants. Based on previous experiments\cite{Vivoni98}, we take $C_N \approx 1$.

For consistency of analysis, we calculate the steady flow as a solution of \eqref{navier-stokes}, and examine its linear stability. The steady uniform flow $U(y)$ resulting from solution of \eqref{navier-stokes} is expected to be unstable, leading to coherent eddies underlying monami. Steady state flow driven by constant pressure gradient satisfies
$-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< H$. 
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=H$) of the fluid layer, as can be seen from the comparison 
between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{fig:basicflow}. Flow consist of a region of approximately constant velocity
\small$U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$\normalsize arising from balance of drag force with pressure gradient with-in grass and a parabolic velocity 
profile in unvegetated region due to balance of viscous force and pressure gradient. Continuity of flow speed and shear stress at canopy top 
give rise to a boundary layer in the flow profile near the tip of the grass. The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this boundary layer thickness is obtained by balance of viscous force with drag, and matching the shear with that in unvegetated region. This leads to a shear layer 
thickness $\delta = \epsilon^{1/3} H$, where $\epsilon = \left[({\rho U_0 H}/\mu) (C_N d N_g H)\right]$, $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity, and
$H$ is half channel width. The dependence of the boundary layer thickness on the grass density $N_g$ gives us a way to systematically control it and investigate its effect on the instability mechanism. 
\begin{figure}[htb]
  \includegraphics[scale=1]{Grass_Base_Vivoni_shear}
\caption{Base flow profile (left )from the experiment of Vivoni\cite{Vivoni98} (Case-3A (Figure-3.6A) with 330 plants/$m^2$, 6 blades per plant, 
plant height = 16.0$\pm0.2$ and blade width of 0.28cm)
 and it's approximation with $U_0=31cm/s$ and $\delta = 0.2H$ with our model. The model of canopy is shown in the bottom and the variation of shear layer in the inset }
\label{fig:basicflow}
\end{figure}

Considering small perturbations $u, v, p$ about the base profile $U$ and $P$ respectively, equation \eqref{navier-stokes} expanded to linear order in the perturbations yields
\begin{equation}
\begin{split}
\frac{\del u}{\del t}+U\frac{\del u}{\del x}+v\frac{\del U}{\del y} &= -\frac{1}{\rho}\frac{\del p}{\del x}+\frac{\mu}{\rho}\nabla^2u-2S(y) C_{N}dN_{g}Uu, \\
\frac{\del v}{\del  t}+ U\frac{\del v}{\del x} &= -\frac{1}{\rho}\frac{\del p}{\del y}+\frac{\mu}{\rho}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0.
 %\nabla\cdot \bu &= 0
\end{split}
\end{equation}
These equations can be non-dimensionalized using half channel height $H$, velocity $U_0$, and the associated advection time $H/U_0$ . 
We are left with three non-dimensional parameters in the model, viz. the Reynolds number $\Rey= {\rho U_0 H}/{\mu}$, the vegetation drag parametrized as $\delta/H$, 
and the submergence ratio of the grass $h_g/H$. With these scaling, and using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a wave solution of the form $\left(u,v,\psi \right)= \left(\hat u, \hat v, \hat\phi \right)e^{ikx+\sigma t}$ to  obtain a modified Orr-Sommerfield equation. The eigenvalue $\sigma$ for a given $k$ provides growth rate and frequency associated with a perturbation of wavelength $2\pi/k$ through the solution of
% \small
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&-D\left(2\epsilon^{-1} S(y) U D \phi\right)
\label{Orr-somerfield}
\end{split}
\end{equation}
\shreyas{Boundary conditions?}
% \normalsize
